Seed ecology of European mesic meadows: the domestication of plant regeneration strategies

### 2021-05-27 doc/manuscript.Rmd
1st full draft 🌱

---
title: "Seed germination ecology of European mesic grasslands"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
    number_sections: false
bibliography: refs.bib
csl: journal-of-vegetation-science.csl
link-citations: yes
always_allow_html: true
---

@@

*Arrhenatheretalia*, *Arrhenatherion*, plant regeneration, seed germination, seed morphology, species-rich meadows, hay meadows, mesic pastures

@@

European mesic grasslands are semi-natural open habitats that occupy moderately fertile and well-drained soils [@RN4698]. These plant communities developed in clearings of a formerly temperate woodland landscape as a consequence of anthropogenic intervention [@RN5033]. At a time, they were novel plant communities with a composite flora made up of species from different origins: grasslands plants that existed in non-forested "islands" of the landscape, plus plants that evolved in parallel to human intervention by hybridisation or polyploidization [@RN5033]. During historical times the botanical composition of mesic grasslands evolved following changes in human activities and management practices [@RN5035]. They are thus the result of a process of habitat domestication that is characteristic of Europe's natural history [@RN5034]. Nowadays, mesic grasslands maintained by agricultural practices are a definitory aspect of European landscapes, both real [@RN4996] and imagined [@RN5029].

Traditional phytosociological classification of mesic grasslands puts emphasis on the separation between mesic meadows used for hay making versus mesic pastures maintained by grazing, but a recent revision at the European level showed that the main driver of variation in species composition is the intensity rather than the type of management [@RN4893]. Indeed, changes in the intensity of management, including land abandonment and agricultural intensification, are threatening the maintenance of mesic grasslands in large parts of Europe [@RN4997]. For this reason the European Habitats Directive (92/43/EEC) has included in its lists of habitats of conservation interest those species-rich meadows that are traditionally managed by one or two annual cuts and light grazing.

One of the reasons for the high conservation interest of traditional mesic grasslands is their high species richness, and for this same reason they have been highlighted as a valuable source of natural seed materials to be used in ecological restoration, rewilding and urban greening [@RN5030; @RN4894; @RN5031]. However, a lack of knowledge about species germination traits has already been identified as a bottleneck [@RN2241] hampering the development of a competitive native seed industry [@RN4665] in Europe. Understanding seed germination is part of the practical scientific framework needed to tackle large-scale ecological restoration challenges [@RN3507] and to maintain *ex situ* collections of plant genetic resources [@RN3763]. When such a knowledge framework is available, ambitious regional schemes of seed-based landscape restoration can be designed [@RN4666].

Mesic grasslands are dominated by mesophilous grasses of the Poaceae family, which make up most of the biomass and define the structure of the vegetation. The dispersal unit in the grasses is generally the floret (**Fig. 1a**), a composite structure made up of the caryopsis or grain (a dry, indehiscent and monospermic fruit) and its surrounding bracts (i.e. modified leaves): the lemma and the palea. Removal of the lemma and palea, as well as puncturing the pericarp, can aid in seed germination [@RN1248]. This structure is often accompanied by awns or hairs (**Fig. 1b**) that are assumed to aid dispersal, although it is not always the case that they do [@RN1290]. Grass seeds have been reported as having physiological seed dormancy [@RN5018; @RN3214; @RN5024] in various degrees, but germinability is usually high even without treating the seeds with cold stratification [@RN1205; @RN1260; @RN1271; @RN1229; @RN372; @RN789; @RN680; @RN1290; @RN398; @RN223; @RN334; @RN1281; @RN5024; @RN3273]. Freshly harvested seeds are comparatively more dormant, but dormancy tends to disappear quickly in dry storage [@RN5024; @RN5018]. Germination has been reported to occur at temperatures ranging from 5 to 30 ºC [@RN3273; @RN1229; @RN1261; @RN1230; @RN1292; @RN5024]. Most Poaceae species have also been reported to germinate better in light than in darkness [@RN1271; @RN1248; @RN1230; @RN1221; @RN1172; @RN1292; @RN1280; @RN5024] and even to be inhibited to germinate by darkness [@RN1229]. Nonetheless, species of *Bromus* have been reported as germinating better in darkness [@RN1172] and *Cynosurus cristatus* as being indifferent to light/darkness [@RN1292; @RN1280]. In *Poa trivialis*, germination was promoted by light but not by alternating temperatures [@RN1205], although alternating temperatures did encourage some germination in darkness [@RN1229]. Wild grass ecotypes usually have a germination response to alternating temperatures [@RN1248; @RN1261; @RN1230; @RN1221; @RN1290; @RN1172; @RN1292; @RN1280], although this response is missing in some wild ecotypes and in the domesticated cultivars [@RN1261], and some species such as *Lolium perenne* have been reported as insensitive to temperature alternation [@RN1280; @RN1380].

Next in abundance to the grasses are the legumes of the Fabaceae family. Legumes contribute to the nutritional value of grassland fodder, as thanks to their N-fixating capabilities they have high N contents [@RN5028; @RN5040]. The dispersal unit of most legumes is the seed itself (**Fig. 1c**) but in some species dispersal units are more complex, including indehiscent monospermic fruits (e.g. *Onobrychis*, **Fig. 1d**) or indehiscent monospermic fruit fragments, i.e. loments (e.g. *Ornithopus*, **Fig. 1e**). Legumes are generally hard-seeded, i.e. they have a water-impermeable seed coat which needs to become permeable before germination can happen (i.e. physical seed dormancy) [@RN3214; @RN3273; @RN1215; @RN5016; @RN3073]. In *Medicago*, seeds that have not reached full maturity can germinate and are not yet impermeable, but the completion of maturation imposed coat impermeability, and thereafter the seed coat must be scarified to allow water imbibition and germination [@RN652]. Buried legume seeds are able to track the seasons, and in some species germination seems to be promoted by cold stratification and alternating temperatures [@RN780], some of these have been described as having combinational dormancy (i.e. physical + physiological) [@RN5037]. However, seeds have also been reported to germinate without any previous treatment [@RN881; @RN632; @RN460; @RN372], and to lose dormancy during storage [@RN5037]. As in the grasses, legume seeds have been reported to germinate in high numbers across a range of temperatures from 5 to 25 ºC [@RN652; @RN3273]. Legume seeds have been described as not responsive to light and capable of germinating in darkness [@RN1339; @RN3273].

Grasses and legumes are accompanied by a diversity of other families which, even if present in lower abundances, contribute to the aesthetic value of the meadows [@RN5038; @RN5039]. They also add nutritional quality, being richer than grasses and legumes in specific elements [@RN5028; @RN5040]. In many of the most frequent minor families (e.g. in the Asteraceae, Ranunculaceae, Polygonaceae, Lamiaceae, Dipsacaceae, Cyperaceae) the dispersal unit is the achene: dry, indehiscent and monospermic fruits. The morphology of these achenes is varied: cypselas with a hairy pappus in *Centaurea* (Asteraceae, **Fig. 1f**); beaked in *Ranunculus* (Ranunculaceae, **Fig. 1g**); hardened nutlets in *Prunella* (Lamiaceae, **Fig. 1h**); hairy in *Knautia* (Dipsacaceae, **Fig. 1i**); surrounded by a perigynium which aids in dispersal by water in *Carex* (Cyperaceae, **Fig. 1j**). In the genus *Sanguisorba* (Rosaceae), the dispersal unit is the urn-shaped receptacle containing one to three achenes (**Fig. 1k**). In the Apiaceae, it is the mericarp (**Fig. 1l**), an indehiscent monospermic fragment of the fruit. In some other minor families the dispersal unit is the seed itself, such as in the Caryophyllaceae (**Fig. 1m**), the Plantaginaceae (**Fig. 1n**), the Juncaceae or the hemiparasitic species of *Rhinanthus* (Orobanchaceae) (**Fig. 1o**). Our understanding of the germination ecology of these seeds is as varied as their morphology. In the Asteraceae, high germination without pretreatments has been reported in *Taraxacum officinale* [@RN1276; @RN194; RN120; @RN963; @RN1428] and *Hypochaeris radicata* [@RN194; @RN1403] and *Achillea millefolium* [@RN1403]. *Taraxacum officinale* germinated between 5 and 30 ºC [@RN120; RN1428; @RN1276], and had higher germination in light [@RN1015; @RN963; @RN1172] and in alternating temperatures [@RN1428]. In *Stachys officinalis* (Lamiaceae), seeds need either cold stratification, light or alternating temperatures to germinate [@RN416; @RN128]. Underdevelopped embryos that need to grow inside the seed before germination (i.e. morphological dormancy) are widespread in the Ranunculaceae and the Apiaceae [@RN3214; @RN1112]. *Ranunculus repens* (Ranunculaceae) germinated between 10 and 25 ºC, but the germination percentages were low [@RN931]; the same species has been reported to respond to alternating temperatures, which can promote its germination even in darkness [@RN3180]. In the Polygonaceae, *Rumex acetosa* can germinate inmediately after dispersal and between 7 and 27 ºC, while the congeneric *Rumex acetosella* does not, this differences is due to the former being able to germinate at constant temperatures in the darkness [@RN814; @RN3273], while the latter has an absolute requirement for light [@RN814]. In *Heracleum sphondylium* (Apiaceae), growth of the embryo only occurs below 10 ºC, in moist conditions [@RN1112]. *Sanguisorba minor* (Rosaceae) increased its germination after abrasion of the seeds with bleach [@RN194; @RN226], although germination without pre-treatment has also been reported [@RN298; @RN226]. Seeds of the hemiparasitic species *Rhinanthus angustifolius* and *Rhinanthus minor* (Orobanchaceae) required relatively long periods of cold stratification to germinate [@RN4014], and were able to germinate in the dark [@RN2236].

Although a wealth of studies has accumulated, a synthesis of the seed germination ecology of mesic grasslands is missing, despite their conservation, restoration and cultural significance. In this article, we present for the first time a synthesis of germination ecology in European mesic grasslands. To do so, we combined a newly-generated dataset with germination records from ENSCOBASE [@RN5036], the seed germination database of the European Native Seed Conservation Network (http://enscobase.maich.gr/index.tml); and the *SylvanSeeds* database of seed germination records for nemoral biome species [@RN4968]. To the resulting dataset containing `r MSrecords` germination records from `r MScountries` European countries we applied Bayesian meta-analysis [@RN5000] to test the seed germination response to environmental cues including scarification, stratification, temperature, alternating temperature and light. In addition, we also analyzed seed morphology (mass and shape) and the differences between the main functional groups of mesic grasslands: grasses (Poaceae), legumes (Fabaceae) and the other families.

@@

## Selection of study species

To create a species list of representative mesic grassland species for inclusion in our analysis, we used a dataset of 118 vegetation relevés from three Western European regions with well-maintained mesic grasslands: 43 from Northern Portugal (Trás-os-Montes), 25 from the Cantabrian Mountains of Spain (Parque Nacional Picos de Europa) and 50 from the Pyrenees. This dataset represents a sample of mesic grassland diversity, as the plots were situated along a major stress gradient related to drought [@RN2980]: two regions with a Temperate macroclimate, and Northern Portugal which is transitional between the Temperate and Mediterranean macroclimates [@RN4893]. Furthermore, grasslands from the Pyrenees are closest to the central European optimum of mesic grassland vegetation (*Arrhenatheretalia elatioris* Tx. 1931) while Portuguese grasslands are in suboptimal areas at the limit of the distribution [@RN4893]. Finally, traditional management of mesic grasslands is relatively well preserved in this area within the European context [@RN4998; @RN4999].

Each relevé had been conducted in a square area (25-100 m^2^ area) situated in the central part of a meadow avoiding the margins. The chosen meadows are maintained by traditional agricultural practices (mowing for hay-making once or twice per year and light grazing). Vegetation in the plots had been sampled in 2016-2017, at the peak of plant development, just before mowing. All vascular plant species in the plots had been recorded and assigned an abundance value using the transformation of the Braun-Blanquet scale to coverage (+ = 0.1%, 1 = 5%, 2 = 17.5%, 3 = 37.5%, 4 = 62.5% and 5 = 87.5%). All plant names in this manuscript follow the nomenclature of the Euro+Med plant database (https://www.emplantbase.org/home.html). Using all the relevés, we calculated the cumulative abundance of each species in the entire area. We removed 208 species with cumulative abundance values below 2%, considering them transient species that might have been recorded by chance and do not represent the core mesic grasslands flora. We used the remaining 117 species as the core list to retrieve seed germination data for this article.

## Seed collection and morphology

From the vegetation plots described above, we collected dispersal units (hereafter called seeds) during the dispersal seasons of 2016, 2017 and 2018. To describe seed morphology, we acquired images of 100-seed samples of each species using a flatbed scanner (Brother LC985) with a resolution of 200 dpi and a scanning area of 1024 x 1024 pixels [@RN2678]. We distributed the seeds on the scanner transparent glass, in a 10 x 10 grid. For each sample, and without moving the seeds, we repeated the scans with black and white backgrounds. In the case of the black background, we covered samples with a black box to avoid interference from environmental light. For the white background, we used the scanner cover. We digitized the obtained images and stored them in JPEG format (Joint Photographic Experts Group). We processed the scanned images using *ImageJ*, an open-source image processing program designed for scientific multidimensional images [@RN4708]. The program calculates several biometric parameters for each seed on the sample, among these, seed length and width. Using these values, we calculated seed shape as seed length divided by seed width. Additionally, we obtained seed mass for the study species from Kew's Seed Information Database [@RN2987].

## Seed germination data

We germinated seeds of the different species using three germination treatments to determine the effect of temperatures representative for the study area: 14/4 ºC representing the capacity of freshly-dispersed seeds to germinate at cool temperatures of spring and autumn, 22/12 ºC as the optimal summer temperature, and 30/20 ºC as sun-heated soil, e.g. soil exposed to sun after hay cutting. Additionally, we compared, at each of these temperature regimes, the germination of fresh seeds versus seeds subjected to a dormancy-breaking treatment. In the case of the Fabaceae, the treatment consisted in scarification by chipping the seed coat with a scalpel, to remove physical dormancy [@RN3214]. For the rest of the families, we used gibberellic acid GA~3~ (0,0645 mM) in darkness during 24h, as a treatment to remove potential physiological seed dormancy [@RN4976]. For each species and treatment, we sowed four Petri dishes with 25 seeds each. The germination substrate was 1% distilled water - agar. We sealed dishes with Parafilm to prevent desiccation. Trials took place in a germination chamber (KBW 400, Binder GmbH, Tuttlingen, Germany) with a 12/12 hour photoperiod (the light period corresponding to the higher temperature). Experiments lasted for four weeks, with germination scoring once per week. The germination criterion was 2 mm radicle emergence. After four weeks, we cut the seeds that failed to germinate and examined them under a magnifying glass. We classified them as normal when the embryo was visible and firm, empty when they lacked an embryo, and contaminated when they were moldy. We only considered normal seeds to calculate germination proportions and conduct subsequent analyses.

In addition to this experimental data, we retrieved seed germination records from ENSCOBASE [@RN5036], the seed germination database of the European Native Seed Conservation Network (http://enscobase.maich.gr/index.tml); and the *SylvanSeeds* database of seed germination records for nemoral biome species [@RN4968]. The combined dataset, including our own experimental data and the records from ENSCOBASE and *SylvanSeeds*, contained `r MSrecords` germination records (i.e. germination proportions for a given seed lot of a species, recorded in a set of laboratory experimental conditions) from `r MScountries` European countries. There were `r MSspeciesN` species in the dataset. Overall, `r MSseedsN` seeds had been used in the experiments. The range of experimental germination temperatures (weighted average of the daily thermoperiod) which had been used in the experiments spans from `r MSminT` to `r MSmaxT` ºC, with `r MSaltN` records of constant temperatures (i.e. experiments that used the same temperature during all their duration) and `r MSaltY` of alternating temperatures (i.e. experiments where different temperatures were applied during the day and the night, in diurnal cycles). Seeds had been exposed to light during some part of the diurnal cycle in `r MSlightY` records or kept in total darkness in `r MSlightN` records. Experiments had been performed with unstratified seeds (i.e. not subjected to a previous dormancy-breaking incubation) in `r MSstratN` records and with stratified seeds (i.e. subjected to previous incubation in dormancy-breaking conditions, including treatments of wet incubation under cold, warm and combinations of cold and warm conditions) in `r MSstratY` records. Finally, there were `r MSgaY` records where GA~3~ had been applied, and `r MSscarY` records where seeds had been scarified.

@@

The mass of these dispersal units in grassland grasses ranges from `r MSpoaMin` to `r MSpoaMax` mg, and they tend to be longer (`r MSpoaLength` mm on average) than they are wide (`r MSpoaWidth` mm).

In mesic grasslands, the dispersal units of the Fabaceae are generally heavier than those of the Poaceae, ranging from `r MSlegMin` to `r MSlegMax` mg; and they are also rounder (`r MSlegLength` mm long vs. `r MSlegWidth` mm wide).

When comparing the morphology of all species (**Fig. 2**), the Poaceae appear as having relatively lighter and more elongated dispersal units, in contrast with the Fabaceae, which tend to have rounder and heavier seeds.

## Germination responses

>>>>

---
title: "Seed ecology of European mesic meadows: the domestication of plant regeneration strategies"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
    number_sections: false
bibliography: refs.bib
csl: journal-of-vegetation-science.csl
link-citations: yes
always_allow_html: true
---

@@

*Arrhenatheretalia*, *Arrhenatherion*, plant regeneration, seed germination, seed morphology, species-rich meadows, hay meadows, mesic grasslands

@@

European mesic meadows are semi-natural open habitats that occupy moderately fertile and well-drained soils [@RN4698]. These plant communities developed in clearings of a temperate woodland landscape as a consequence of anthropogenic intervention [@RN5033; @RN5045]. At a time, they were novel plant communities with a composite flora made up of species from different origins: grasslands plants that existed in non-forested "islands" of the landscape, plus plants that evolved in parallel to human intervention by hybridization or polyploidization [@RN5033]. During historical times the botanical composition of mesic meadows evolved following changes in human activities and management practices [@RN5035; @RN5045]. They are thus the result of a process of habitat domestication that is characteristic of Europe's natural history [@RN5034]. Nowadays, mesic meadows maintained by agricultural practices are an essential aspect of European landscapes, both real [@RN4996] and imagined [@RN5029].

Traditional phytosociological classification of mesic grasslands puts emphasis on the separation between meadows used for hay making versus pastures maintained by grazing, but a recent revision at the European level showed that the main driver of variation in species composition is the intensity rather than the type of management [@RN4893]. Indeed, changes in the intensity of management, including land abandonment and agricultural intensification, are threatening the maintenance of mesic meadows in large parts of Europe [@RN4997]. For this reason, the European Habitats Directive (92/43/EEC) has included in its lists of habitats of conservation interest those species-rich meadows that are traditionally managed by one or two annual cuts and light grazing.

One of the reasons for the high conservation interest of traditional mesic meadows is their high species richness, and for this same reason they have been highlighted as a valuable source of natural seed materials to be used in ecological restoration, rewilding and urban greening [@RN5030; @RN4894; @RN5031]. However, a lack of knowledge about species germination traits has already been identified as a bottleneck [@RN2241] hampering the development of a competitive native seed industry [@RN4665] in Europe. Understanding seed germination is part of the practical scientific framework needed to tackle large-scale ecological restoration challenges [@RN3507] and to maintain *ex situ* collections of plant genetic resources [@RN3763]. When such a knowledge framework is available, ambitious regional schemes of seed-based landscape restoration can be designed [@RN4666].

Mesic meadows are dominated by mesophilous grasses of the Poaceae family, which make up most of the biomass and define the structure of the vegetation. The dispersal unit in the grasses is generally the floret (**Fig. 1a**), a composite structure made up of the caryopsis or grain (a dry, indehiscent and monospermic fruit) and its surrounding bracts (i.e. modified leaves): the lemma and the palea. Removal of the lemma and palea, as well as puncturing the pericarp, can aid in seed germination [@RN1248]. The dispersal unit is often accompanied by awns or hairs (**Fig. 1b**) that are assumed to aid dispersal, although it is not always the case that they do [@RN1290]. Meadow grass seeds have been reported as having physiological seed dormancy [@RN5018; @RN3214; @RN5024] in various degrees, but germinability is usually high even without treating the seeds with cold stratification [@RN1205; @RN1260; @RN1271; @RN1229; @RN372; @RN789; @RN680; @RN1290; @RN398; @RN223; @RN334; @RN1281; @RN5024; @RN3273]. Freshly harvested seeds are comparatively more dormant, but dormancy tends to disappear quickly in dry storage [@RN5024; @RN5018]. Germination has been reported to occur at temperatures ranging from 5 to 30 ºC [@RN3273; @RN1229; @RN1261; @RN1230; @RN1292; @RN5024]. Most Poaceae species have also been reported to germinate better in light than in darkness [@RN1271; @RN1248; @RN1230; @RN1221; @RN1172; @RN1292; @RN1280; @RN5024] and even to be inhibited to germinate by darkness [@RN1229]. Nonetheless, species of *Bromus* have been reported as germinating better in darkness [@RN1172] and *Cynosurus cristatus* as being indifferent to light/darkness [@RN1292; @RN1280]. In *Poa trivialis*, germination was promoted by light but not by alternating temperatures [@RN1205], although alternating temperatures did encourage some germination in darkness [@RN1229]. Wild grass ecotypes usually have a germination response to alternating temperatures [@RN1248; @RN1261; @RN1230; @RN1221; @RN1290; @RN1172; @RN1292; @RN1280], although this response is missing in some wild ecotypes and in the domesticated cultivars [@RN1261], and some species such as *Lolium perenne* have been reported as insensitive to temperature alternation [@RN1280; @RN1380].

Next in abundance to the grasses are the legumes of the Fabaceae family. Legumes contribute to the nutritional value of meadow fodder, as thanks to their N-fixating capabilities they have high N contents [@RN5028; @RN5040]. The dispersal unit of most legumes is the seed itself (**Fig. 1c**) but in some species dispersal units are more complex, including indehiscent monospermic fruits (e.g. *Onobrychis*, **Fig. 1d**) or indehiscent monospermic fruit fragments, i.e. loments (e.g. *Ornithopus*, **Fig. 1e**). Legume seeds are generally hard-seeded, i.e. they have a water-impermeable seed coat which needs to become permeable before germination can happen (i.e. physical seed dormancy) [@RN3214; @RN3273; @RN1215; @RN5016; @RN3073]. In *Medicago*, seeds that have not reached full maturity can germinate before they become impermeable, but the completion of maturation imposes coat impermeability, and thereafter the seed must be scarified to allow water imbibition and germination [@RN652]. Buried legume seeds are able to track the seasons, and in some species germination seems to be promoted by cold stratification and alternating temperatures [@RN780]; some of these species have been described as having combinational dormancy (i.e. physical + physiological) [@RN5037]. However, seeds have also been reported to germinate without any previous treatment [@RN881; @RN632; @RN460; @RN372], and to lose dormancy during storage [@RN5037]. As in the grasses, legume seeds have been reported to germinate in high numbers across a range of temperatures from 5 to 25 ºC [@RN652; @RN3273]. Legume seeds have been described as not responsive to light and capable of germinating in darkness [@RN1339; @RN3273].

Grasses and legumes are accompanied by a diversity of other families which, even if present in lower abundances, contribute to the aesthetic value of the meadows [@RN5038; @RN5039]. They also add nutritional scope, being richer than grasses and legumes in specific elements [@RN5028; @RN5040]. In many of the most frequent minor families (e.g. in the Asteraceae, Ranunculaceae, Polygonaceae, Lamiaceae, Dipsacaceae, Cyperaceae) the dispersal unit is the achene: dry, indehiscent and monospermic fruits. The morphology of these achenes is varied: cypselae with a hairy pappus in *Centaurea* (Asteraceae, **Fig. 1f**); beaked in *Ranunculus* (Ranunculaceae, **Fig. 1g**); hardened nutlets in *Prunella* (Lamiaceae, **Fig. 1h**); hairy in *Knautia* (Dipsacaceae, **Fig. 1i**); surrounded by a perigynium which aids in dispersal by water in *Carex* (Cyperaceae, **Fig. 1j**). In the genus *Sanguisorba* (Rosaceae), the dispersal unit is the urn-shaped receptacle containing one to three achenes (**Fig. 1k**). In the Apiaceae, it is the mericarp (**Fig. 1l**), an indehiscent monospermic fragment of the fruit. In some other minor families the dispersal unit is the seed itself, such as in the Caryophyllaceae (**Fig. 1m**), the Plantaginaceae (**Fig. 1n**), the Juncaceae or the hemiparasitic species of *Rhinanthus* (Orobanchaceae) (**Fig. 1o**). In the Asteraceae, high germination without pretreatments has been reported in *Taraxacum officinale* [@RN1276; @RN194; @RN120; @RN963; @RN1428], *Hypochaeris radicata* [@RN194; @RN1403] and *Achillea millefolium* [@RN1403]. *Taraxacum officinale* germinates between 5 and 30 ºC [@RN120; RN1428; @RN1276], and has higher germination in light [@RN1015; @RN963; @RN1172] and in alternating temperatures [@RN1428]. In *Stachys officinalis* (Lamiaceae), seeds need either cold stratification, light or alternating temperatures to germinate [@RN416; @RN128]. Underdeveloped embryos that need to grow inside the seed before germination (i.e. morphological dormancy) are widespread in the Ranunculaceae and the Apiaceae [@RN3214; @RN1112]. *Ranunculus repens* (Ranunculaceae) germinates between 10 and 25 ºC, but the germination percentages were reported to be low [@RN931]; the same species has been reported to respond to alternating temperatures, which can promote its germination even in darkness [@RN3180]. In the Polygonaceae, *Rumex acetosa* can germinate immediately after dispersal and between 7 and 27 ºC, while the congeneric *Rumex acetosella* does not, this difference is due to the former being able to germinate at constant temperatures in the darkness [@RN814; @RN3273], while the latter has an absolute requirement for light [@RN814]. In *Heracleum sphondylium* (Apiaceae), growth of the embryo only occurs below 10 ºC, in moist conditions [@RN1112]. *Sanguisorba minor* (Rosaceae) increases its germination after abrasion of the seeds with bleach [@RN194; @RN226], although germination without pre-treatment has also been reported [@RN298; @RN226]. Seeds of the hemiparasitic species *Rhinanthus angustifolius* and *Rhinanthus minor* (Orobanchaceae) require relatively long periods of cold stratification to germinate [@RN4014], and are able to germinate in the dark [@RN2236].

Although a wealth of studies has accumulated, a synthesis of the seed ecology of mesic meadows is missing, despite their conservation, restoration and cultural significance. In this article, we present for the first time a synthesis of regeneration by seed in European mesic meadows. To do so, we combine newly-generated data on seed morphology and germination with records from databases [@RN5036; @RN4968; @RN2987; @RN2319]. To the resulting dataset containing `r MSrecords` germination records from `r MScountries` European countries we apply Bayesian meta-analysis [@RN5000] to test the seed germination response to environmental cues including scarification, stratification, temperature, alternating temperature and light. We analyze the relationship between seed traits and species environmental preferences, and compare the germination ecology of mesic meadows with that of other herbaceous plant communities.

@@

## Selection of mesic meadow species

To create a list of representative mesic meadow species for inclusion in our analysis, we used a dataset comprising 118 vegetation relevés (i.e. records of plants species co-occurring in sampling plots) from three Western European regions with well-maintained mesic meadows: 43 relevés from Northern Portugal, 25 from the Cantabrian Mountains of Spain and 50 from the Pyrenees. These relevés contain a sample of mesic meadow diversity, as they were recorded along a major stress gradient related to summer drought [@RN2980]: the Pyrenees and Cantabrian Mountains have a temperate macroclimate, whereas Northern Portugal is transitional between the temperate and Mediterranean macroclimates. Furthermore, meadows from the Pyrenees are closest to the European optimum of mesic meadow vegetation while the Portuguese ones are in suboptimal areas at the limit of the European distribution of temperate meadows [@RN4893]. Finally, the traditional management of meadows is relatively well preserved in these three regions compared to their European context [@RN4998; @RN4999].

Sampled meadows had been maintained by traditional agricultural practices: mowing for hay-making once or twice per year plus light grazing. Each sampling plot was placed in a square area (25-100 m^2^ area) situated in the central part of a meadow, avoiding the margins. Sampling took place in 2016-2017, at the peak of plant development, just before mowing. All vascular plant species in the plots were recorded and given a cover value using the transformation of the Braun-Blanquet scale to coverage (+ = 0.1%, 1 = 5%, 2 = 17.5%, 3 = 37.5%, 4 = 62.5% and 5 = 87.5%). All plant names were assigned following the nomenclature of Euro+Med [-@RN5044], which is used throughout this article. As expected, the vegetation of the sampled meadows was dominated by the Poaceae and the Fabaceae: these two families represented 47% and 17%, respectively, of the total plant cover recorded in all the plots. Other 20 families were recorded, the largest of which was the Asteraceae; each of these other families represented less than 10% of the total cover. These numbers refer only to the seed plants, pteridophytes being a marginal part of the records.

Using all the relevés, we calculated the cumulative cover of each species in the entire area. To perform the calculation, first we standardized the cover values of the plots by dividing the cover of each species in each plot by the total plant cover in that plot. Then, for each species, we calculated its total cover in the dataset, by summing its standardized cover values from all the plots. Finally, we rescaled the values of all species to a 1-100 scale to obtain the cumulative cover values. We removed 208 species with cumulative cover values below 2%, considering them to be transient species that might have been recorded by chance and may not represent the core mesic meadows flora. We used the remaining 116 species as the core list to retrieve seed germination data for this article. The majority of these species were hemicryptophytes (78%), with some therophytes (16%) and a few chamaephytes and geophytes (3% of each). The family with more species in the list was the Poaceae (22%), with another 17% belonging to the Fabaceae, 15% to the Asteraceae, 8% to the Apiaceae, and the rest of the families representing less than 5% each.

We also used the relevés as a basis to characterize the preferences of the selected species for three ecological drivers of mesic meadow diversity: cold, summer drought and soil reaction [@RN2980]. For cold and drought, we used the coordinates of the plots to retrieve from CHELSA [@RN4766] the bioclimatic variables bio06 (minimum temperature of the coldest month) and bio14 (precipitation of the driest month). and represents the limiting water availability along the Iberian gradient of summer drought. For soil reaction, we took from each plot five soil samples from between 0 and 20 cm depth with a Dutch auger and combined them to make a bulk soil horizon, which we subsequently air-dried, crumbled, finely crushed and sieved with a 2 mm screen, to finally measure the pH in H~2~O with a glass electrode in a suspension of soil and water (1:2.5). For each of these three environmental variables (bio06, bio14, pH) measured at the plot level we calculated the species niche centroids (SNCs). The SNC for any given species and variable is the mean of the environmental variable in all the plots were the species occurs, weighted by species cover in each plot [@RN4709]. The list of core species, with their cumulative covers and SNCs, is available at GitHub (see Data Availability Statement).

## Seed morphology and germination dataset

From the vegetation plots described above, we collected dispersal units (hereafter called seeds) during the dispersal seasons of 2016, 2017 and 2018. Seed collection followed the methodology of ENSCONET [-@RN2907]. To describe seed morphology, we acquired images of 100-seed samples of each species using a flatbed scanner (Brother LC985) with a resolution of 200 dpi and a scanning area of 1024 x 1024 pixels [@RN2678]. We distributed the seeds on the transparent glass of the scanner, in a 10 x 10 grid. For each sample, and without moving the seeds, we repeated the scans with black and white backgrounds. In the case of the black background, we covered samples with a black box to avoid interference from environmental light. For the white background, we used the scanner cover. We digitized the obtained images and stored them in JPEG format (Joint Photographic Experts Group). We processed the scanned images using *ImageJ*, an open-source image processing program designed for scientific multidimensional images [@RN4708]. The program calculates several biometric parameters for each seed on the sample, and among these we choose seed length and width. Additionally, we retrieved species values of seed mass from the Seed Information Database [@RN2987] and of seed number at the individual/ramet level from the LEDA database [@RN2319]. The dataset with the length and width measures is available at GitHub (see Data Availability Statement).

We germinated the collected seeds using three germination treatments to determine the germination response to temperatures that are representative of the study area: 14/4 ºC representing the capacity of freshly-dispersed seeds to germinate at cool temperatures of spring and autumn, 22/12 ºC as the summer temperature, and 30/20 ºC as sun-heated soil, e.g. soil exposed to sun after hay cutting. Additionally, we compared, at each of these temperature regimes, the germination of fresh seeds versus seeds subjected to a dormancy-breaking treatment. In the case of the Fabaceae and other families that might present physical dormancy [@RN3214], the treatment consisted in scarification by chipping the seed coat with a scalpel. For the rest of the families, we used gibberellic acid GA~3~ (0,0645 mM) in darkness during 24h, as a treatment to remove potential physiological seed dormancy [@RN4976]. For each species and treatment, we sowed four Petri dishes with 25 seeds each. The germination substrate was 1% distilled water - agar. We sealed dishes with Parafilm to prevent desiccation. Trials took place in a germination chamber (KBW 400, Binder GmbH, Tuttlingen, Germany) with a 12/12-hour photoperiod (the light period corresponding to the higher temperature). Experiments lasted for four weeks, with germination scoring once per week. The germination criterion was 2 mm radicle emergence. After four weeks, we cut the seeds that failed to germinate and examined them under a magnifying glass. We classified them as normal when the embryo was visible and firm, empty when they lacked an embryo, and contaminated when they were moldy. We only considered normal seeds to calculate germination proportions and conduct subsequent analyses.

In addition to this experimental data, we retrieved seed germination records from ENSCOBASE [@RN5036], the seed germination database of the European Native Seed Conservation Network (http://enscobase.maich.gr/index.tml); and the *SylvanSeeds* database of seed germination records for the nemoral biome [@RN4968]. The combined dataset, including our own experimental data and the records from ENSCOBASE and *SylvanSeeds*, contained `r MSrecords` germination records (i.e. germination proportions for a given seed lot of a species, recorded in a set of laboratory experimental conditions) from `r MScountries` European countries. There were `r MSspeciesN` species in the dataset (i.e. 90% of the core list of meadow species was covered). Overall, `r MSseedsN` seeds had been used in the experiments. The range of experimental germination temperatures (weighted average of the daily thermoperiod) which had been used in the experiments spanned from `r MSminT` to `r MSmaxT` ºC, with `r MSaltN` records of constant temperatures (i.e. experiments that used the same temperature during all their duration) and `r MSaltY` of alternating temperatures (i.e. experiments where different temperatures were applied during the day and the night, in diurnal cycles). Seeds had been exposed to light during some part of the diurnal cycle in `r MSlightY` records or kept in total darkness in `r MSlightN` records. Experiments had been performed with unstratified seeds (i.e. not subjected to a previous dormancy-breaking incubation) in `r MSstratN` records and with stratified seeds (i.e. subjected to previous incubation in dormancy-breaking conditions, including treatments of wet incubation under cold, warm and combinations of cold and warm conditions) in `r MSstratY` records. Finally, there were `r MSgaY` records where GA~3~ had been applied, and `r MSscarY` records where seeds had been scarified. The full germination dataset is available at GitHub (see Data Availability Statement).

@@

The Poaceae had lower values of seed mass and higher values of seed number, while the Fabaceae had heavier but fewer seeds (**Fig. 2**). The other families covered the range of values showed by the Poaceae and the Fabaceae, but their median values were high for both traits: their median seed mass was close to that of the Fabaceae, while their median seed number was higher than that of the Poaceae (**Fig. 2**). Seed shape also showed a divergence between the Poaceae and Fabaceae, with seeds being elongated in the former and rounder in the latter (**Fig. 2**). The other families covered the full range of variation, with both elongated and round seeds (**Fig. 2**).

## Seed germination

When considering the full set of mesic meadow species, all of the six studied germination cues had a significant effect on final germination proportions (**Fig. 3**). The germination of meadow seeds was positively associated with scarification, stratification, GA~3~, alternating temperatures, and light. Average temperature had a negative effect, indicating a trend towards higher germination at lower temperatures. Averaging the whole dataset, the highest germination proportions were achieved at 20 ºC. Between 0 ºC and 20 ºC, germination proportions increased steadily with increasing temperatures. Above 20 ºC, germination declined more sharply, and meadow seeds rarely germinated at 30 ºC.
